---
title: "Coordination Game Analysis"
subtitle: "Common Ground × Advice Type Experiment"
author: "Robert & Yuka"
date: "2026-01-30"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
---

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(lme4)
library(emmeans)
library(knitr)
library(here)

# Set theme for all plots
theme_set(theme_minimal(base_size = 14))
```

## Overview

This analysis examines how common ground (CG) and advice type (cooperative vs. selfish) affect coordination in a repeated parking game. Participants were paired and played 13 rounds, choosing parking spots to maximize earnings while avoiding collisions.

- **Common Ground**: Both players know they received the same advice (CG) vs. only know their own advice (noCG)
- **Advice Type**: "Try to find an outcome that's fair for both of you" (cooperative) vs. "Try to get the best outcome for yourself" (selfish)

## Data Loading

```{r load-data, message=FALSE}
# Load data
player <- read.csv(here("data_analysis/data_from_jan27_full/combined_player.csv"))
playerRound <- read.csv(here("data_analysis/data_from_jan27_full/combined_playerRound.csv"))
games_for_analysis <- read.csv(here("data_analysis/data_from_jan27_full/games_for_analysis.csv"))

# Get treatment info and filter to analyzed games (passed attention checks)
treatment_info <- player %>% select(gameID, treatmentName) %>% distinct()

player_filtered <- player %>%
  filter(gameID %in% games_for_analysis$id) %>%
  mutate(
    isCG = as.numeric(grepl("^CG_", treatmentName)),
    isCooperate = as.numeric(grepl("_cooperate$", treatmentName))
  )

cat("Total players:", nrow(player_filtered), "\n")
cat("Total games:", length(unique(player_filtered$gameID)), "\n")
```

```{r treatment-counts}
# Treatment distribution
kable(table(player_filtered$treatmentName),
      col.names = c("Condition", "N Players"),
      caption = "Sample size by condition")
```

## Game-Level Analysis

For game level analysis we aggregate to one observation per pair since both players in a game share outcomes.

```{r game-level-data}
player_filtered
game_level <- player_filtered %>%
  group_by(gameID, treatmentName, isCG, isCooperate) %>%
  summarize(dif_bonus = abs(diff(bonus)),bonus = mean(bonus), .groups = "drop")

cat("Games per condition:\n")cat("Games perdiff() condition:\n")
print(table(game_level$treatmentName))
```

### Main Effects Model

```{r game-level-models}
# Model comparisons
base_model <- lm(bonus ~ 1, data = game_level)
cg_model <- lm(bonus ~ isCG, data = game_level)
independent_model <- lm(bonus ~ isCG + isCooperate, data = game_level)
interaction_model <- lm(bonus ~ isCG * isCooperate, data = game_level)
```

```{r cg-effect}

kable(anova(base_model, cg_model), digits = 3)
kable(anova(base_model, independent_model), digits = 3)
kable(anova(independent_model, interaction_model), digits = 3)
```

```{r game-level-dif-models}
# Model comparisons
dif_base_model <- lm(dif_bonus ~ 1, data = game_level)
dif_cg_model <- lm(dif_bonus ~ isCG, data = game_level)
dif_independent_model <- lm(dif_bonus ~ isCG + isCooperate, data = game_level)
dif_interaction_model <- lm(dif_bonus ~ isCG * isCooperate, data = game_level)
```

```{r cg-effect}

kable(anova(dif_base_model, dif_cg_model), digits = 3)
kable(anova(dif_base_model, dif_independent_model), digits = 3)
kable(anova(dif_independent_model, dif_interaction_model), digits = 3)
```

Common ground significantly improves coordination across the board (p = 0.047). No other effects.

## Round-Level Analysis

Now we examine dynamics over the course of the 13 rounds.

```{r round-level-data}
# Prepare round-level data
playerRound_proc <- playerRound %>%
  left_join(treatment_info, by = "gameID") %>%
  filter(gameID %in% games_for_analysis$id) %>%
  arrange(gameID, playerID, decisionLastChangedAt) %>%
  group_by(playerID) %>%
  mutate(round = row_number() - 1) %>%
  ungroup() %>%
  mutate(
    isCG = as.numeric(grepl("^CG_", treatmentName)),
    isCooperate = as.numeric(grepl("_cooperate$", treatmentName)),
    cum_bonus = bonus
  )
for (i in 0:11) {
  playerRound_proc$cum_bonus[playerRound_proc$round == i+1] <- 
    playerRound_proc$bonus[playerRound_proc$round == i+1] + 
    playerRound_proc$cum_bonus[playerRound_proc$round == i]
}
# Aggregate to game-round level
game_round <- playerRound_proc %>%
  group_by(gameID, round, treatmentName, isCG, isCooperate) %>%
  summarize(
    p1_choice = decision[1],
    p2_choice = decision[2],
    p1_bonus = bonus[1],
    p2_bonus = bonus[2],
    p1_cum_bonus = cum_bonus[1],
    p2_cum_bonus = cum_bonus[2],
    .groups = "drop"
  ) %>%
  mutate(
    collision = as.numeric(p1_choice == p2_choice),
    coordinated = 1 - collision,
    round_bonus = p1_bonus + p2_bonus,
    bonus_diff = abs(p1_bonus - p2_bonus),
    cum_bonus_diff = abs(p1_cum_bonus-p2_cum_bonus),
    CG = ifelse(isCG == 1, "Common Ground", "No Common Ground"),
    Advice = ifelse(isCooperate == 1, "Cooperative", "Selfish"),
    A = (p1_choice == "A") + (p2_choice == "A"),
    B = (p1_choice == "B") + (p2_choice == "B"),
    C = (p1_choice == "C") + (p2_choice == "C"),
    D = (p1_choice == "D") + (p2_choice == "D")
  )
game_round_before <- game_round%>%mutate(
  round_before = round+1,
  prev_p1_choice = p1_choice,
  prev_p2_choice=p2_choice,
)
game_round %>% left_join(game_round_before%>%select(gameID,round_before,prev_p1_choice,prev_p2_choice), by = c('gameID','round'='round_before'))
cat("Round-level observations:", nrow(game_round), "\n")
```

### Collision Rates by Condition

```{r collision-summary}
collision_summary <- game_round %>%
  group_by(treatmentName) %>%
  summarize(
    `N Rounds` = n(),
    `Collision Rate` = sprintf("%.1f%%", mean(collision) * 100),
    `Mean Pair Bonus` = round(mean(round_bonus), 1),
    `Mean Bonus Diff` = round(mean(bonus_diff), 2),
    .groups = "drop"
  )

kable(collision_summary, caption = "Summary statistics by condition")
```

### Learning Over Time

```{r learning-summary}
learning <- game_round %>%
  mutate(phase = ifelse(round < 6, "Early (0-5)", "Late (6-12)")) %>%
  group_by(treatmentName, phase) %>%
  summarize(collision_rate = mean(collision) * 100, .groups = "drop") %>%
  tidyr::pivot_wider(names_from = phase, values_from = collision_rate) %>%
  mutate(Improvement = `Early (0-5)` - `Late (6-12)`) %>%
  arrange(desc(Improvement))

kable(learning, digits = 1,
      caption = "Collision rates: early vs late rounds (% points improvement)")
```

CG + Selfish shows the biggest learning effect (11.8 percentage points improvement).

### Model 1: Coordination (Binary)

```{r model-coordination, warning=FALSE}
m_coord <- glmer(coordinated ~ isCG * isCooperate * scale(round) + (1 | gameID),
                  data = game_round, family = binomial,
                  control = glmerControl(optimizer = "bobyqa"))

kable(summary(m_coord)$coefficients, digits = 3,
      caption = "Coordination model (logistic mixed effects)")
```

Significant 3-way interaction (p = 0.029): Learning trajectories differ by condition.

### Model 2: Fairness (Bonus Difference)

```{r model-fairness}
m_fair <- lmer(bonus_diff ~ isCG * isCooperate * scale(round) + (1 | gameID),
               data = game_round)

kable(summary(m_fair)$coefficients, digits = 3,
      caption = "Fairness model (linear mixed effects)")
m_fair <- lmer(cum_bonus_diff ~ isCG * isCooperate * scale(round) + (1 | gameID),
               data = game_round)

kable(summary(m_fair)$coefficients, digits = 3,
      caption = "Fairness model (linear mixed effects)")
```

3-way interaction (p = 0.004): CG + Selfish becomes increasingly unfair over time.

## Viz

### Bonus Over Time

```{r plot-bonus, fig.height=5, fig.width=10, message=FALSE}
ggplot(game_round, aes(x = round, y = round_bonus, color = Advice, fill = Advice)) +
  geom_smooth(method = "lm", span = 0.75, alpha = 0.2) +
  facet_wrap(~CG) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_color_manual(values = c("Cooperative" = "#2E86AB", "Selfish" = "#A23B72")) +
  scale_fill_manual(values = c("Cooperative" = "#2E86AB", "Selfish" = "#A23B72")) +
  labs(x = "Round", y = "Pair Bonus",
       title = "Coordination Bonus Over Time",
       subtitle = "Loess smoothing with 95% CI") +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```

### Fairness Over Time

```{r plot-fairness, fig.width=10, fig.height=5, message=FALSE}
ggplot(game_round, aes(x = round, y = bonus_diff, color = Advice, fill = Advice)) +
  geom_smooth(method = "lm", alpha = 0.2) +
  facet_wrap(~CG) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_color_manual(values = c("Cooperative" = "#2E86AB", "Selfish" = "#A23B72")) +
  scale_fill_manual(values = c("Cooperative" = "#2E86AB", "Selfish" = "#A23B72")) +
  labs(x = "Round", y = "Bonus Difference Between Partners",
       title = "Fairness Over Time",
       subtitle = "Lower = more equal outcomes (linear fit with 95% CI)") +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```
### Cumulative Bonus difference

```{r plot-fairness, fig.width=10, fig.height=5, message=FALSE}
ggplot(game_round, aes(x = round, y = cum_bonus_diff, color = Advice, fill = Advice)) +
  geom_smooth(method = "lm", alpha = 0.2) +
  facet_wrap(~CG) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_color_manual(values = c("Cooperative" = "#2E86AB", "Selfish" = "#A23B72")) +
  scale_fill_manual(values = c("Cooperative" = "#2E86AB", "Selfish" = "#A23B72")) +
  labs(x = "Round", y = "Total Bonus Difference Between Partners",
       title = "Cumulative Bonus Difference",
       subtitle = "Lower = more equal outcomes (linear fit with 95% CI)") +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```



### Probablity of decisions

```{r plot-bonus, fig.height=5, fig.width=10, message=FALSE}
ggplot(game_round, aes(x = round, y = C/2, color = Advice, fill = Advice)) +
  geom_smooth(method = "lm", span = 0.75, alpha = 0.2) +
  facet_wrap(~CG) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_color_manual(values = c("Cooperative" = "#2E86AB", "Selfish" = "#A23B72")) +
  scale_fill_manual(values = c("Cooperative" = "#2E86AB", "Selfish" = "#A23B72")) +
  labs(x = "Round", y = "Probability choosing C",
       title = "Choice of C Over Time",
       subtitle = "Loss smoothing with 95% CI") +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```