---
author: "Robert & Yuka"
date: "2026-01-30"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    title: Stats for Cogsci
---

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(lme4)
library(emmeans)
library(knitr)
library(here)
library(brms)
library(jsonlite)
library(stringr)
library(lmerTest)
# Set theme for all plots
theme_set(theme_minimal(base_size = 14))
```

## Overview

This analysis contains all code for producing graphs and stats in the Cogsci 20226 paper, in which we examine how common ground (CG) and advice type (cooperative vs. selfish) affect coordination in a repeated parking game. Participants were paired and played 13 rounds, choosing parking spots to maximize earnings while avoiding collisions.

-   **Common Ground**: Both players know they received the same advice (CG) vs. only know their own advice (noCG)
-   **Advice Type**: "Try to find an outcome that's fair for both of you" (cooperative) vs. "Try to get the best outcome for yourself" (selfish)

All codes for cogsci 2026 Common Ground Affects Coordination paper:

```{r load-data, message=FALSE}
# Load data
player <- read.csv(here("data_analysis/data_from_jan27_full/combined_player.csv"))
playerRound <- read.csv(here("data_analysis/data_from_jan27_full/combined_playerRound.csv"))
games_for_analysis <- read.csv(here("data_analysis/data_from_jan27_full/games_for_analysis.csv"))

# Get treatment info and filter to analyzed games (passed attention checks)
treatment_info <- player %>% select(gameID, treatmentName) %>% distinct()

player_filtered <- player %>%
  filter(gameID %in% games_for_analysis$id) %>%
  mutate(
    isCG = as.numeric(grepl("^CG_", treatmentName)),
    isCooperate = as.numeric(grepl("_cooperate$", treatmentName))
  )

cat("Total players:", nrow(player_filtered), "\n")
cat("Total games:", length(unique(player_filtered$gameID)), "\n")

```
## Analyze Actions across contexts
### Format data
```{r format-data}
player_filtered
game_level <- player_filtered %>%
  group_by(gameID, treatmentName, isCG, isCooperate) %>%
  summarize(dif_bonus = abs(diff(bonus)),bonus = mean(bonus), .groups = "drop")

cat("Games per condition:\n")
print(table(game_level$treatmentName))
# Prepare round-level data
playerRound_proc <- playerRound %>%
  left_join(treatment_info, by = "gameID") %>%
  filter(gameID %in% games_for_analysis$id) %>%
  arrange(gameID, playerID, decisionLastChangedAt) %>%
  group_by(playerID) %>%
  mutate(round = row_number() - 1) %>%
  ungroup() %>%
  mutate(
    isCG = as.numeric(grepl("^CG_", treatmentName)),
    isCooperate = as.numeric(grepl("_cooperate$", treatmentName)),
    cum_bonus = bonus,
    C=ifelse(decision=="C",1,0)
  )
for (i in 0:11) {
  playerRound_proc$cum_bonus[playerRound_proc$round == i+1] <- 
    playerRound_proc$bonus[playerRound_proc$round == i+1] + 
    playerRound_proc$cum_bonus[playerRound_proc$round == i]
}

# Aggregate to game-round level
game_round <- playerRound_proc %>%
  group_by(gameID, round, treatmentName, isCG, isCooperate) %>%
  summarize(
    p1_choice = decision[1],
    p2_choice = decision[2],
    p1_bonus = bonus[1],
    p2_bonus = bonus[2],
    p1_cum_bonus = cum_bonus[1],
    p2_cum_bonus = cum_bonus[2],
    .groups = "drop"
  ) %>%
  mutate(
    collision = as.numeric(p1_choice == p2_choice),
    coordinated = 1 - collision,
    round_bonus = p1_bonus + p2_bonus,
    bonus_diff = abs(p1_bonus - p2_bonus),
    cum_bonus_diff = abs(p1_cum_bonus-p2_cum_bonus),
    CG = ifelse(isCG == 1, "Common Ground", "No Common Ground"),
    Advice = ifelse(isCooperate == 1, "Cooperative", "Selfish"),
    A = (p1_choice == "A") + (p2_choice == "A"),
    B = (p1_choice == "B") + (p2_choice == "B"),
    C = (p1_choice == "C") + (p2_choice == "C"),
    D = (p1_choice == "D") + (p2_choice == "D")
  )
game_round_before <- game_round%>%mutate(
  round_before = round+1,
  prev_p1_choice = p1_choice,
  prev_p2_choice=p2_choice,
)
game_round <-game_round %>% left_join(game_round_before%>%select(gameID,round_before,prev_p1_choice,prev_p2_choice), by = c('gameID','round'='round_before'))

cat("Round-level observations:", nrow(game_round), "\n")
```

### Game-level Analysis
```{r}
#  CG condition coordinated better (currently significance is weak)
bonus_model_base <- lm(bonus ~  1,game_level)
bonus_model <- lm(bonus ~ isCG,game_level)
anova(bonus_model,bonus_model_base)

# There were no significant collision differences across conditions
C_model <- lm(collision ~ 1,game_round%>%group_by(gameID)%>%summarize(collision=sum(collision),isCooperate=isCooperate[1]))
C_model_interact <- lm(collision~ isCooperate,game_round%>%group_by(gameID)%>%summarize(collision=sum(collision),isCooperate=isCooperate[1]))
anova(C_model,C_model_interact)

# But Participants in selfish condition chose C more (weak significance)
C_model <- lm(C ~ 1,playerRound_proc%>%group_by(gameID)%>%summarize(C=sum(C)))
C_model_interact <- lm(C ~ isCooperate,playerRound_proc%>%group_by(gameID)%>%summarize(C=sum(C),isCooperate=isCooperate[1]))
anova(C_model,C_model_interact)
```
### Round-Level Analysis

Now we examine dynamics over the course of the 13 rounds.

```{r example-trace}
# 01KG0H3X9S8ZR8HTJY0HSJCWW9

```

```{r reported-stats}
# Participants' coordination improved as rounds progressed
round_model_null<- lmer(round_bonus ~ 1+ (1|gameID),game_round)
round_model <- lmer(round_bonus ~ round + (1|gameID),game_round)
anova(round_model,round_model_null)
```

#### Analyzing round-level bonus differences
```{r round-level-bonus, fig.width=10, fig.height=5, message=FALSE}
m_round_dif <- lmer(bonus_diff ~ isCG * isCooperate * scale(round) + (1 | gameID),
               data = game_round)

kable(summary(m_round_dif)$coefficients, digits = 3,
      caption = "round-level dif model (linear mixed effects)")
ggplot(game_round, aes(x = round, y = bonus_diff, color = Advice, fill = Advice)) +
  geom_smooth(method = "lm", alpha = 0.2) +
  facet_wrap(~CG) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_color_manual(values = c("Cooperative" = "#2E86AB", "Selfish" = "#A23B72")) +
  scale_fill_manual(values = c("Cooperative" = "#2E86AB", "Selfish" = "#A23B72")) +
  labs(x = "Round", y = "Bonus Difference Between Partners",
       title = "Fairness Over Time",
       subtitle = "Lower = more equal outcomes (linear fit with 95% CI)") +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```
Round level, diff, but this doesn't translate to cumulative diff:

#### Analyzing cumulative bonus differences
```{r cumulative-bonus}
m_fair <- lmer(cum_bonus_diff ~ isCG * isCooperate * scale(round) + (1 | gameID),
               data = game_round)

kable(summary(m_fair)$coefficients, digits = 3,
      caption = "Fairness model (linear mixed effects)")

ggplot(game_round, aes(x = round, y = cum_bonus_diff, color = Advice, fill = Advice))+
  geom_smooth(method = "lm", alpha = 0.2) +
  facet_wrap(~CG) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_color_manual(values = c("Cooperative" = "#2E86AB", "Selfish" = "#A23B72")) +
  scale_fill_manual(values = c("Cooperative" = "#2E86AB", "Selfish" = "#A23B72")) +
  labs(x = "Round", y = "Total Bonus Difference Between Partners",
       title = "Cumulative Bonus Difference",
       subtitle = "Lower = more equal outcomes (linear fit with 95% CI)") +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))


```
(CG doesn't have any effect on fairness!)

so people must be compensating across rounds by swappinng - but swaps are too few frequency to get significant results:
#### Analyzing swaps (Need Edits!)
```{r model-alt-purple}
library(lmerTest)
game_round <- game_round%>%
  rowwise()%>%
  mutate(
  alt_purple = ifelse((C==1)&(D==1)&(prev_p1_choice==p2_choice)&(prev_p2_choice==p1_choice),1,0)
)
summary_group <- game_round[game_round$round>8,]%>%group_by(CG,Advice)%>%summarize(alt_purple=mean(alt_purple),.groups = "drop")
kable(summary_group,caption = "Difference in alt strategy across conditions")

alt_model<-glmer(alt_purple ~isCG*isCooperate + (1|gameID),data=game_round[game_round$round>5,],family = binomial)
kable(summary(alt_model)$coefficients, digits = 3,
      caption = "Effects of Condition on alt strategy in later rounds (6-12) - very sensitive to round numbers so not trustworthy")

ggplot(game_round%>%filter(round>0), aes(x = round, y = alt_purple, color = Advice, fill = Advice)) +
  stat_summary(fun = mean, geom = "point", size = 2) +  # Add this
  geom_smooth(method = "lm", span = 0.75, alpha = 0.2) +
  facet_wrap(~CG) +
  scale_x_continuous(breaks = seq(1, 12, 2)) +
  scale_color_manual(values = c("Cooperative" = "#2E86AB", "Selfish" = "#A23B72")) +
  scale_fill_manual(values = c("Cooperative" = "#2E86AB", "Selfish" = "#A23B72")) +
  labs(x = "Round", y = "Pair Bonus",
       title = "Frequency of swaps overtime",
       subtitle = "Loess smoothing with 95% CI") +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```

## Analyse "We" in written responses

```{r we-data-prep}
# prepare data
player_writing <- player_filtered %>%
  rowwise() %>%
  summarise(
    CG = fromJSON(treatment)$adviceType, 
    Cooperate = fromJSON(treatment)$advice,
    adviceFollowed = fromJSON(exitSurvey)$adviceFollowed,
    partnerAdvice = fromJSON(exitSurvey)$partnerAdvice,
    writtenPlan = fromJSON(writtenPlan)$plan,
    we_used = str_count(tolower(adviceFollowed), "we\\b"),
    I_used = str_count(tolower(adviceFollowed), "\\bi\\b"),
    we_at_start = str_count(tolower(writtenPlan), "we\\b"),
    bonus = bonus,
    gameID = gameID
  )
games_writing <- player_writing%>%group_by(gameID) %>%
  summarize(dif_bonus = abs(bonus[1]-bonus[2]), we_used = we_used[1]+we_used[2], we_at_start = we_at_start[1]+we_at_start[2], I_used = sum(I_used),Cooperate = Cooperate[1],CG=CG[1])
```

### uses of we in written plan
```{r we-at-start-analyses}
# Q. does advice type change the way we think about the game? Does it matter whether the advice is in common ground?
advice_on_perception <- lm(we_at_start ~ Cooperate*CG, data = games_writing)
summary(advice_on_perception)
kable(games_writing%>%group_by(Cooperate,CG)%>%summarise(we_at_start=mean(we_at_start)))
# A. Advice type does , but CG is very weak (p = approx. 0.2).

# Q. does our initial perception predict fairness in the game?
perception_on_fairness <- lm(dif_bonus ~ we_at_start*CG, data = games_writing)
summary(perception_on_fairness)
# A. weakly (p=0.126). And there is some interaction between we at start and CG - we at start affects fairness if advice is in common ground bur not if not in common ground.
```

### uses of we in reflection (only CG condition)
```{r we-at-end-analyses}
# Q. Does fairness affect joint intention (perception of coordination in the game)
cg_writing <- games_writing%>%filter(CG=="CG")
model <- lm(dif_bonus ~ we_used, data =cg_writing)
summary(model)
# A. yes. p=0.02 (I guess this is the opposite, not we_used ~ dif_bonus)
ggplot(cg_writing, aes(x = we_used, y = dif_bonus)) +
  geom_jitter(width = 0.05, height = 0, alpha = 0.6, size = 2)+
  geom_smooth(method = "lm",alpha=0.2, color = "blue", fill = "lightblue") +
  labs(title = "Relationship between 'We' Usage in reflection, and Bonus Difference",
       x = "Number of 'We' Used",
       y = "Bonus Difference") +
  theme_minimal()
```

```{r I-at-end-analyses}
model <- lm(dif_bonus ~ I_used, data =cg_writing)
summary(model)
# A. yes. p=0.02 (I guess this is the opposite, not we_used ~ dif_bonus)
ggplot(cg_writing, aes(x = I_used, y = dif_bonus)) +
  geom_jitter(width = 0.05, height = 0, alpha = 0.6, size = 2)+
  geom_smooth(method = "lm",alpha=0.2, color = "blue", fill = "lightblue") +
  labs(title = "Relationship between 'I' Usage in reflection, and Bonus Difference",
       x = "Number of 'I' Used",
       y = "Bonus Difference") +
  theme_minimal()
```