---
author: "Robert & Yuka"
date: "2026-01-30"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    title: Stats for Cogsci
---

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(lme4)
library(emmeans)
library(knitr)
library(here)
library(brms)
library(jsonlite)
library(stringr)
library(lmerTest)
# Set theme for all plots
theme_set(theme_minimal(base_size = 14))
```

## Overview

This analysis contains all code for producing graphs and stats in the Cogsci 20226 paper, in which we examine how common ground (CG) and advice type (cooperative vs. selfish) affect coordination in a repeated parking game. Participants were paired and played 13 rounds, choosing parking spots to maximize earnings while avoiding collisions.

-   **Common Ground**: Both players know they received the same advice (CG) vs. only know their own advice (noCG)
-   **Advice Type**: "Try to find an outcome that's fair for both of you" (cooperative) vs. "Try to get the best outcome for yourself" (selfish)

All codes for cogsci 2026 Common Ground Affects Coordination paper:

```{r load-data, message=FALSE}
# Load data
player <- read.csv(here("data_analysis/data_from_jan27_full/combined_player.csv"))
playerRound <- read.csv(here("data_analysis/data_from_jan27_full/combined_playerRound.csv"))
games_for_analysis <- read.csv(here("data_analysis/data_from_jan27_full/games_for_analysis.csv"))
strategies <- read.csv(here("data_analysis/data_from_jan27_full/classified_strategies.csv"))

# Get treatment info and filter to analyzed games (passed attention checks)
treatment_info <- player %>% select(gameID, treatmentName) %>% distinct()

player_filtered <- player %>%
  filter(gameID %in% games_for_analysis$id) %>%
  mutate(
    isCG = as.numeric(grepl("^CG_", treatmentName)),
    isCooperate = as.numeric(grepl("_cooperate$", treatmentName))
  )

cat("Total players:", nrow(player_filtered), "\n")
cat("Total games:", length(unique(player_filtered$gameID)), "\n")

```
## Analyze Actions across contexts
### Format data
```{r format-data}
game_level <- player_filtered %>%
  group_by(gameID, treatmentName, isCG, isCooperate) %>%
  summarize(dif_bonus = abs(diff(bonus)),bonus = mean(bonus), .groups = "drop")

cat("Games per condition:\n")
print(table(game_level$treatmentName))
# Prepare round-level data
playerRound_proc <- playerRound %>%
  left_join(treatment_info, by = "gameID") %>%
  filter(gameID %in% games_for_analysis$id) %>%
  arrange(gameID, playerID, decisionLastChangedAt) %>%
  group_by(playerID) %>%
  mutate(round = row_number() - 1) %>%
  ungroup() %>%
  mutate(
    isCG = as.numeric(grepl("^CG_", treatmentName)),
    isCooperate = as.numeric(grepl("_cooperate$", treatmentName)),
    cum_bonus = bonus,
    C=ifelse(decision=="C",1,0)
  )
for (i in 0:11) {
  playerRound_proc$cum_bonus[playerRound_proc$round == i+1] <- 
    playerRound_proc$bonus[playerRound_proc$round == i+1] + 
    playerRound_proc$cum_bonus[playerRound_proc$round == i]
}

# Aggregate to game-round level
game_round <- playerRound_proc %>%
  group_by(gameID, round, treatmentName, isCG, isCooperate) %>%
  summarize(
    p1_choice = decision[1],
    p2_choice = decision[2],
    p1_bonus = bonus[1],
    p2_bonus = bonus[2],
    p1_cum_bonus = cum_bonus[1],
    p2_cum_bonus = cum_bonus[2],
    .groups = "drop"
  ) %>%
  mutate(
    collision = as.numeric(p1_choice == p2_choice),
    coordinated = 1 - collision,
    round_bonus = p1_bonus + p2_bonus,
    bonus_diff = abs(p1_bonus - p2_bonus),
    CG = ifelse(isCG == 1, "Common Ground", "No Common Ground"),
    Advice = ifelse(isCooperate == 1, "Cooperative", "Selfish"),
    A = (p1_choice == "A") + (p2_choice == "A"),
    B = (p1_choice == "B") + (p2_choice == "B"),
    C = (p1_choice == "C") + (p2_choice == "C"),
    D = (p1_choice == "D") + (p2_choice == "D")
  )
game_round_before <- game_round%>%mutate(
  round_before = round+1,
  prev_p1_choice = p1_choice,
  prev_p2_choice=p2_choice,
)
game_round <-game_round %>% left_join(game_round_before%>%select(gameID,round_before,prev_p1_choice,prev_p2_choice), by = c('gameID','round'='round_before'))

cat("Round-level observations:", nrow(game_round), "\n")
```

### Low-resolution Analysis
```{r}
#  CG condition coordinated better
bonus_model <- lm(bonus ~  isCG+isCooperate,game_level)
summary(bonus_model)

# No difference in fairness
dif_bonus_model <- lm(dif_bonus ~ isCG+isCooperate,game_level)
summary(dif_bonus_model)

# There were no significant collision differences across conditions
collision_model <- glmer(collision ~isCG+isCooperate + (1|gameID),game_round,family = binomial)
summary(collision_model)

# But Participants in selfish condition chose C more
C_model <- glmer(C ~ isCG+isCooperate+round + (1|playerID),data=playerRound_proc,family = binomial)
summary(C_model)$coefficients
playerRound_proc<-playerRound_proc%>%mutate(early_round = round<)
# This effect was particularly strong in the earlier rounds when participants were establishing norms
C_model_start <- glmer(C ~ isCG*early_round+isCooperate*early_round+(1|playerID),data=playerRound_proc,family = binomial)
summary(C_model_start)$coefficients
```
report p,(z or t),beta for each of them

```{r reported-stats}
# Participants' coordination improved as rounds progressed
round_model_null<- lmer(round_bonus ~ 1+ (1|gameID),game_round)
round_model <- lmer(round_bonus ~ round + (1|gameID),game_round)
summary(round_model)$coefficients
```
report  beta, t and p

### CG+Selfish converged to a different norm
```{r round-level-bonus, fig.width=10, fig.height=5, message=FALSE}
m_round_dif <- lmer(bonus_diff ~ isCG * isCooperate * scale(round) + (1+scale(round) | gameID),
               data = game_round)

kable(summary(m_round_dif)$coefficients, digits = 3,
      caption = "round-level dif model (linear mixed effects)")

ggplot(game_round, aes(x = round, y = bonus_diff, color = Advice, fill = Advice)) +
  geom_smooth(method = "lm", alpha = 0.2) +
  facet_wrap(~CG) +
  # Top label - only on first facet
  annotate("text", x = 0, y = Inf, label = "More Asymmetric", 
           hjust = 0, vjust = 1.5, size = 3.5, fontface = "italic", color = "gray30") +
  # Bottom label - only on first facet
  annotate("text", x = 0, y = -Inf, label = "More Symmetric", 
           hjust = 0, vjust = -0.5, size = 3.5, fontface = "italic", color = "gray30") +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_color_manual(values = c("Cooperative" = "#4CAF50", "Selfish" = "#D32F2F")) +
  scale_fill_manual(values = c("Cooperative" = "#4CAF50", "Selfish" = "#D32F2F")) +
  labs(x = "Round Number", y = "Payoff Difference Between Partners") +theme_minimal()+
  theme(legend.position = c(0.9, 0.9), strip.text = element_text(face = "bold"))
```
Stats to report: p=0.03, t=-2.17, beta=-1.12

Confirm finding from rond-level difference by looking at frequency of strategies in classification
### Alt purple
```{r model-alt-purple}
strategies <- read.csv(here("data_analysis/data_from_jan27_full/classified_strategies.csv"))
strats <- setdiff(colnames(strategies), c("treatmentName", "game_number"))
strategies <- strategies %>%
  mutate(across(all_of(strats), ~ as.numeric(.) * 4))
strategies$CG <- sub("_.*", "", strategies$treatmentName)
strategies$Advice <- sub(".*_", "", strategies$treatmentName)
model <- glm(cbind(alt_purple, 4 - alt_purple) ~ Advice * CG,
             data = strategies,
             family = binomial)

pairs(emmeans(model,~ Advice*CG))
summary(model)
strategies%>%group_by(treatmentName)%>%summarise(alt_purple=mean(alt_purple)/4)
```
stats to report: 8% of last 4 rounds were alt_strategy in CG_selfish, but others were below 1.5%
stats to report: Adviceselfish has a coefficient of 1.86 (z=-0.012, p=.016) which is significanly different from 0 (I'm not sure how to interpret this to say this explains how only CG+Selfish condition went up in round level difference)


### purple equilibrium vs orange equilibrium
```{r which-equilibrium}
model <- glm(cbind(alt_purple+stable_purple, 4 - stable_purple-alt_purple) ~ Advice * CG,
             data = strategies,
             family = binomial)

pairs(emmeans(model,~ Advice * CG))
summary(model)
model <- glm(cbind(stable_orange, 4 - stable_orange) ~ Advice * CG,
             data = strategies,
             family = binomial)

pairs(emmeans(model,~ Advice * CG))
summary(model)
model <- glm(cbind(other, 4 - other) ~ Advice * CG,
             data = strategies,
             family = binomial)

pairs(emmeans(model,~ Advice * CG))
summary(model)
```
Advice and CG both had effect on the type of strategy
For stable_orange, both CG, advice type, and interaction was all significant (reprot betha, z, and p values), and for strategies converging on the purple equilibrium, advice type was significant.

## Analyse "We" in written responses

```{r we-data-prep}
# prepare data
player_writing <- player_filtered %>%
  rowwise() %>%
  summarise(
    isCG = fromJSON(treatment)$adviceType=="CG", 
    isCooperate = fromJSON(treatment)$advice!="Try to get the best outcome for yourself.",
    adviceFollowed = fromJSON(exitSurvey)$adviceFollowed,
    partnerAdvice = fromJSON(exitSurvey)$partnerAdvice,
    writtenPlan = fromJSON(writtenPlan)$plan,
    we_used = str_count(tolower(adviceFollowed), "\\bwe\\b"),
    I_used = str_count(tolower(adviceFollowed), "\\bi\\b"),
    we_at_start = str_count(tolower(writtenPlan), "\\bwe\\b"),
    bonus = bonus,
    gameID = gameID
  )
games_writing <- player_writing%>%group_by(gameID) %>%
  summarize(tot_bonus = sum(bonus),dif_bonus = abs(bonus[1]-bonus[2]), we_used = we_used[1]+we_used[2], we_at_start = we_at_start[1]+we_at_start[2], I_used = sum(I_used),isCooperate = isCooperate[1],isCG=isCG[1])
```

### uses of we in written plan
```{r we-at-start-analyses}
# Q. does advice type change the way we think about the game? Does it matter whether the advice is in common ground?
advice_on_perception_inter <- glm(we_at_start ~ isCooperate*, data = games_writing,family=poisson)
anova(advice_on_perception_inter,advice_on_perception)
summary(advice_on_perception_inter)
kable(games_writing%>%group_by(Cooperate,CG)%>%summarise(we_at_start=mean(we_at_start)))
# A. Advice type does , but CG is very weak (p = approx. 0.2).

# Q. does our initial perception predict fairness in the game?
perception_on_fairness <- lm(dif_bonus ~ we_at_start, data = games_writing)
summary(perception_on_fairness)
# A. weakly (p=0.126). And there is some interaction between we at start and CG - we at start affects fairness if advice is in common ground bur not if not in common ground.
```

### uses of we in reflection (only CG condition)
```{r we-at-end-analyses}
model_base <- lm(dif_bonus ~ we_at_start, data =cg_writing)
summary(glm(we_at_start ~ isCooperate, data =cg_writing,family=poisson))

summary(glm(we_used ~ isCooperate, data =cg_writing,family=poisson))
# Q. Does fairness affect joint intention (perception of coordination in the game)
cg_writing <- games_writing%>%filter(isCG)
model <- lm(dif_bonus ~ we_used+isCooperate, data =cg_writing)
model_base <- lm(dif_bonus ~ we_used, data =cg_writing)
anova(model_base,model)
summary(model_base)
# A. yes. p=0.02 (I guess this is the opposite, not we_used ~ dif_bonus)
summary(glm(we_used~dif_bonus,data=cg_writing,family=poisson))
summary(glm(we_used~dif_bonus+isCooperate,data=cg_writing,family=poisson))
summary(glm(we_used~dif_bonus+isCooperate,data=cg_writing,family=poisson))
ggplot(cg_writing, aes(x = dif_bonus, y = we_used)) +
  geom_jitter(width = 0.05, height = 0, alpha = 0.6, size = 2)+
  geom_smooth(method = "lm",alpha=0.2, color = "blue", fill = "lightblue") +
  labs(title = "Relationship between 'We' Usage in reflection, and Bonus Difference",
       y = "Number of 'We' Used",
       x = "Bonus Difference") +
  theme_minimal()
```

```{r I-at-end-analyses}
model <- lm(dif_bonus ~ I_used+isCooperate, data =cg_writing)
model_indep <- lm(dif_bonus ~ I_used, data =cg_writing)
colnames(cg_writing)
anova(model,model_indep)
cor_result <- cor.test(cg_writing$we_used, cg_writing$I_used)
print(cor_result)
summary(model)
# A. yes. p=0.02 (I guess this is the opposite, not we_used ~ dif_bonus)
ggplot(cg_writing, aes(x = I_used, y = dif_bonus)) +
  geom_jitter(width = 0.05, height = 0, alpha = 0.6, size = 2)+
  geom_smooth(method = "lm",alpha=0.2, color = "blue", fill = "lightblue") +
  labs(title = "Relationship between 'I' Usage in reflection, and Bonus Difference",
       x = "Number of 'I' Used",
       y = "Bonus Difference") +
  theme_minimal()
```






```{r}
library(ggplot2)
library(patchwork)  # or use gridExtra

# Create a list to store plots
plot_list <- list()
experimental_d <- read.csv(here("data_analysis/data_from_jan27_full/classified_strategies.csv"))
for (tName in unique(experimental_d$treatmentName)) {
  experimental_d_sub <- experimental_d[experimental_d$treatmentName == tName, ]
  count = c(sum(experimental_d_sub$stable_orange), sum(experimental_d_sub$stable_purple), 
              sum(experimental_d_sub$alt_purple), sum(experimental_d_sub$both_stable),
              sum(experimental_d_sub$other))
  total <- sum(count)
  frequencies <- count / total
  strategy_counts_exp <- data.frame(
    strategy = c("stable_orange", "stable_purple", "alt_purple", "both_stable","other"),
    frequency = frequencies,
    treatmentName = tName
  )
  
  p_experimental <- ggplot(strategy_counts_exp, aes(x=strategy, y=frequency)) + 
    geom_col(fill = ifelse(grepl("selfish", tName, ignore.case = TRUE), "#D32F2F", "#4CAF50")) +
    labs(x = NULL,y=NULL,title = gsub("_", " + ", tName)) + 
    theme_classic() + 
    scale_x_discrete(limits = c("stable_orange", "stable_purple", "alt_purple", "both_stable","other"),
                     labels = c("stable\norange", "stable\npurple", "alt\npurple", "other\nnorme","no\nnorm")) +
    scale_y_continuous(limits = c(0, 0.65))
  plot_list[[tName]] <- p_experimental
}

# Combine into 2x2 grid using patchwork
combined_plot <- (plot_list[[1]] | plot_list[[2]]) / (plot_list[[4]] | plot_list[[3]])
combined_plot
```